An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects\\ in Time-Frequency Representations

### 2020-07-22 Notes/BoundEffRed.tex

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{euscript}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
%\usepackage{subfig}
\usepackage{color}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{dsfont}
\usepackage[shortlabels]{enumitem}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{hyperref}
\usepackage{mathpazo}
\usepackage{array}
\usepackage{bigints}

@@

\title{An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects in Time-Frequency Representations}

@@

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\includegraphics[width=.24\textwidth]{zoomSSTintro.eps}
\includegraphics[width=.24\textwidth]{zoomSSTBoundEffRed.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~sec.}
\label{fig:ex.intro}
\end{figure}

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. For instance, judicious choices of analysis windows whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane~\cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship of chosen analysis windows and TF analysis tool, these techniques do not make it possible to reduce the boundary effects of any TF representations, and for any analysis window. Another natural idea consists of carrying out a preliminary step of extending the signal beyond its boundaries, and due to its flexibility, various extension schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. There also exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD), the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR), the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}, and long short-term memory model \cite{vlachas2018data}. It is far from an exhaustive list, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. While the second class of extension schemes gives better-extended signals than the first class, they generally have a great computational cost.

In this paper, we propose a fast extension algorithm based on a simple dynamical model, in order to optimize the trade-off between the extension quality and the computational cost. The proposed algorithm is then composed of two steps.
\begin{enumerate}
\item \emph{Extend the signal by forecasting it.} The aim is to use a dynamic model to predict the values taken by the measured signal outside the measurement interval. Then, once this operation is done, we have access to an extended signal defined on a larger interval $I_\Delta$, where $\Delta$ denotes the size of the extension on both boundaries of $I$.
\item \emph{Run the local analysis tool on the extended signal.} Assuming that the support of the analysis window is smaller than $2\Delta$, the local analysis near the boundary of $I$ is now possible without lack of information thanks to knowledge brought by the extension. 
\end{enumerate}
Thus, assuming that the quality of the extension step is sufficient, the analysis results obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the non-extended signal. We claim and prove that forecasting oscillatory signals based on the simple dynamical model combined with the simple least square approach is sufficient for reducing the boundary effects for the TF analysis, or other kernel-based analysis. See the bottom right of Fig.~\ref{fig:ex.intro} for a snapshot of the result. The main benefit of this simple approach is a numerically efficient solution with a theoretical guarantee for real-time analysis purposes.

@@

\section{Theoretical performance}
\label{se:theoretical}
\input{theory}

\section{Numerical results}
\label{se:results}
\input{results}

>>>>

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{euscript}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{color}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{dsfont}
\usepackage[shortlabels]{enumitem}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{hyperref}
\usepackage{mathpazo}
\usepackage{array}
\usepackage{bigints}

@@

\title{An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects\\ in Time-Frequency Representations}

@@

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\includegraphics[width=.24\textwidth]{zoomSSTintro.eps}
\includegraphics[width=.24\textwidth]{zoomSSTBoundEffRed.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~seconds.}
\label{fig:ex.intro}
\end{figure}

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. For instance, judicious choices of analysis windows whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane~\cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship of chosen analysis windows and TF analysis tool, these techniques do not make it possible to reduce the boundary effects of any TF representations, and for any analysis window. Another natural idea consists of carrying out a preliminary step of extending the signal beyond its boundaries, and due to its flexibility, various extension schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. There also exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD), the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR), the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}, and long short-term memory model \cite{vlachas2018data}. It is far from an exhaustive list, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. Whereas the second class of extension schemes gives better-extended signals than the first class, they generally have a great computational cost.

In this paper, we propose a fast extension algorithm based on a simple dynamical model, in order to optimize the trade-off between the extension quality and the computational cost. The proposed algorithm is then composed of two steps.
\begin{enumerate}
\item \emph{Extend the signal by forecasting it.} The aim is to use a dynamic model to predict the values taken by the measured signal outside the measurement interval. Then, once this operation is done, we have access to an extended signal defined on a larger interval $I_\Delta$, where $\Delta$ denotes the size of the extension on both boundaries of $I$.
\item \emph{Run the local analysis tool on the extended signal.} Assuming that the support of the analysis window is smaller than $2\Delta$, the local analysis near the boundary of $I$ is now possible without lack of information thanks to knowledge brought by the extension. 
\end{enumerate}
Thus, assuming that the quality of the extension step is sufficient, the analysis results obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the nonextended signal. We claim and prove that forecasting oscillatory signals based on the simple dynamical model combined with the simple least square approach is sufficient for reducing the boundary effects for the TF analysis, or other kernel-based analysis. See the bottom right of Fig.~\ref{fig:ex.intro} for a snapshot of the result. The main benefit of this simple approach is a numerically efficient solution with a theoretical guarantee for real-time analysis purposes.

@@

\section{Theoretical Performance}
\label{se:theoretical}
\input{theory}

\section{Numerical Results}
\label{se:results}
\input{results}
